---
title: "Nephrology Cohort Analysis-Ready Dataset, from Cerner HealthFacts"
author: "Jeremy Yang"
output:
  html_document:
    number_sections: yes
    toc: yes
---

# Introduction

A cohort of patients has been identified by searching based on several
ICD-9 and ICD-10 diagnostic codes, where "kidney" is in the `diagnosis_description`.
A random 1% sample of data from 2013-2017 results in 75894 `patient_sk`'s and
80459 `patient_id`'s.  From these patients we extract encounter data for 
specific fact types:

* `DIAGNOSIS`
* `LAB_PROCEDURE`
* `MEDICATION`

Not obtained but also available:

* `CLINICAL_EVENT`
* `IMPLANT_LOG`
* `MED_HISTORY`
* `MICROBIOLOGY`
* `MICRO_SUSCEPTIBILITY`
* `PROCEDURE`
* `SURGICAL_CASE`
* `SURGICAL_PROCEDURE`
* `DISCHARGE`

```{r echo=FALSE}
knitr::opts_chunk$set(echo=FALSE, message=FALSE)
base::date()
```

```{r message=FALSE, warning=FALSE}
library(readr, quietly=T)
library(data.table, quietly=T)
library(plotly, quietly=T)
```

# Read files


```{r}
cohort <- read_delim("data/hf_neph_cohort.tsv", "\t", col_types=cols(.default=col_character(), age_in_years=col_integer()))
setDT(cohort)
knitr::kable(cohort[, .(.N), by="gender"])
knitr::kable(cohort[, .(.N), by="race"])
```

```{r}
plot_ly(type="histogram", data=cohort, x=~age_in_years) %>%
  layout(title="Patient Age")
```

```{r}
dia <- read_delim("data/hf_neph_cohort_f_diagnosis.tsv", ",", col_types=cols(.default=col_character()))
lab <- read_delim("data/hf_neph_cohort_f_lab.tsv", ",", col_types=cols(.default=col_character()))
med <- read_delim("data/hf_neph_cohort_f_medication.tsv", ",", col_types=cols(.default=col_character()))
```

