---
title: "Kidney Disease Cohort Analysis-Ready Dataset, from Cerner HealthFacts"
author: "Jeremy Yang"
output:
  html_document:
    number_sections: yes
    toc: yes
---

# Introduction

A cohort of patients has been identified by searching based on several
ICD-9 and ICD-10 diagnostic codes, where "kidney" is in the `diagnosis_description`.
A random 1% sample of data from 2013-2017 results in 75894 `patient_sk`'s and
80459 `patient_id`'s.  From these patients we extract encounter data for 
specific fact types:

* `DIAGNOSIS`
* `LAB_PROCEDURE`
* `MEDICATION`

Not obtained but also available:

* `CLINICAL_EVENT`
* `IMPLANT_LOG`
* `MED_HISTORY`
* `MICROBIOLOGY`
* `MICRO_SUSCEPTIBILITY`
* `PROCEDURE`
* `SURGICAL_CASE`
* `SURGICAL_PROCEDURE`
* `DISCHARGE`

```{r echo=FALSE}
knitr::opts_chunk$set(echo=FALSE, message=FALSE)
base::date()
```

```{r message=FALSE, warning=FALSE}
library(readr, quietly=T)
library(data.table, quietly=T)
library(plotly, quietly=T)
```

# Cohort selection and description

## Exclusions:
- ICD9: E* = Injury to kidney
- ICD9: O* = Tuberculosis or Syphillis of kidney
- ICD9: V* = Family History or Kidney Transplant
- ICD9: 866.* = Injury to Kidney
- ICD9: 966.81 = Complications of Transplanted Kidney
- ICD10-CM: E* = Diabetes with kidney complications
- ICD10-CM: O* = Pregnancy-related kidney complications
- ICD10-CM: R* = Radiologic findings
- ICD10-CM: S* = Injury to kidney
- ICD10-CM: T* = Complication following kidney dialysis or transplant
- ICD10-CM: Y* = Complications of dialysis, etc.
- ICD10-CM: Z* = Complications of transplant, etc.

### SQL:

```
SELECT
  dp.patient_id,
  dp.patient_sk,
  dp.gender,
  dp.race,
  fe.age_in_years,
  fe.patient_type_id,
  fe.hospital_id,
  dd.diagnosis_code,
  dd.diagnosis_type,
  fe.admitted_dt_tm AS dx_date
FROM
  hf_d_patient dp
JOIN
  hf_f_encounter fe ON dp.patient_id = fe.patient_id
JOIN
  hf_f_diagnosis fd1 ON fd1.encounter_id = fe.encounter_id
JOIN
  hf_d_diagnosis dd ON fd1.diagnosis_id = dd.diagnosis_id
WHERE
  EXTRACT(YEAR FROM fe.admitted_dt_tm) BETWEEN 2013 AND 2017
  AND dd.diagnosis_id IN
  ( SELECT
    diagnosis_id
  FROM
    hf_d_diagnosis
  WHERE
    dd.diagnosis_description ILIKE '%kidney%'
    AND NOT (dd.diagnosis_type = 'ICD9' AND (dd.diagnosis_code ~ '^[EOV]' OR dd.diagnosis_code ~ '^(866|996)'))
    AND NOT (dd.diagnosis_type = 'ICD10-CM' AND dd.diagnosis_code ~ '^[EORSTYZ]')
  )
  AND RANDOM() < 0.01
  ;
```

```{r warning=FALSE}
icd_neph <- read_delim("data/hf_neph_codes.tsv", "\t", col_types=cols(.default=col_character()))
setDT(icd_neph)
cohort <- read_delim("data/hf_neph_cohort.tsv", "\t", col_types=cols(.default=col_character(), age_in_years=col_integer()))
setDT(cohort)
message(sprintf("Total (1%% sample): %d; Estimated full total: %g", nrow(cohort), 100*nrow(cohort)))
```

## Cohort by gender

```{r}
knitr::kable(cohort[, .(.N, Percent=round(100.0*.N/nrow(cohort), 2)), by="gender"], caption="Cohort by gender")
```

## Cohort by race

```{r}
ax0 <- list(showline=F, zeroline=F, showticklabels=F, showgrid=F)
plot_ly(type="pie", hole = 0.5, data=cohort[, .(.N), by="race"], values=~N, labels=~race ) %>%
  layout(title=paste0("Cohort by race"), 
         xaxis=ax0, yaxis=ax0, showlegend=T, margin=list(t=120), legend=list(x=0.8, y=0.5),
         font=list(family="monospace", size=18))
```

## Cohort age

```{r}
plot_ly(type="histogram", data=cohort[!is.na(age_in_years)], x=~age_in_years)  %>%
  layout(title=paste0("Cohort age in years"), 
         showlegend=F, margin=list(t=120),
         font=list(family="monospace", size=18))
```


## Qualifying diagnoses

The dx facts file contains ___all___ dx for the cohort, but first consider their qualifying
kidney disease dx.

### SQL:

```
SELECT DISTINCT
  dd.diagnosis_id,
  dd.diagnosis_code,
  dd.diagnosis_type,
  dd.diagnosis_description,
  fe.encounter_id,
  fe.patient_id,
  fe.patient_type_id,
  fe.age_in_years,
  fe.hospital_id,
  fe.admitted_dt_tm AS date
FROM
  hf_f_diagnosis fd
JOIN
  hf_f_encounter fe ON fd.encounter_id = fe.encounter_id
JOIN
  hf_d_diagnosis dd ON fd.diagnosis_id = dd.diagnosis_id
JOIN
  hf_d_diagnosis_type ddt ON fd.diagnosis_type_id = ddt.diagnosis_type_id
WHERE
  fe.patient_id IN (...)
  ;
```

```{r}
dia <- read_delim("data/hf_neph_cohort_f_diagnosis.tsv", ",", col_types=cols(.default=col_character()))
setDT(dia)
dia[, `:=`(Ftype=NULL, Result=NULL, Units=NULL, Page=NULL, PtypeID=NULL)]
setnames(dia, old="FID", new="diagnosis_id")
message(sprintf("Percent of SKs in Dx facts file: %.1f%%", 100*uniqueN(dia[, SK])/uniqueN(cohort[, patient_sk])))
dx_counts <- dia[, .(N_dx = .N, N_pt=uniqueN(SK)), by="diagnosis_id"]
dx_counts <- merge(dx_counts, icd_neph, by="diagnosis_id", all.x=F, all.y=T)
set(dx_counts, which(is.na(dx_counts[, N_dx])), "N_dx", value=0)
setorder(dx_counts, -N_dx)
knitr::kable(dx_counts[1:20, .(Dx=paste(diagnosis_type, diagnosis_code, ":"), diagnosis_description, N_dx, N_pt)])
```


## Co-morbidity

Other dx. Some may be criteria for exclusion.

```{r}
icd <- read_delim("data/hf_d_diagnosis.tsv", "\t", col_types=cols(.default=col_character()))
setDT(icd)
codx_counts <- dia[, .(N_dx = .N, N_pt=uniqueN(SK)), by="diagnosis_id"]
codx_counts <- merge(codx_counts, icd, by="diagnosis_id", all.x=F, all.y=T)
codx_counts <- codx_counts[!(diagnosis_id %in% icd_neph$diagnosis_id)]
set(codx_counts, which(is.na(codx_counts[, N_dx])), "N_dx", value=0)
setorder(codx_counts, -N_dx)
knitr::kable(codx_counts[1:20, .(Dx=paste(diagnosis_type, diagnosis_code, ":"), diagnosis_description, N_dx, N_pt)])
```

## Facts: Lab Procedures

```{r warning=FALSE}
lab <- read_delim("data/hf_neph_cohort_f_lab.tsv", ",", col_types=cols(.default=col_character(), Date=col_date()))
setnames(lab, old="FID", new="lab_procedure_id")
setDT(lab)
lab[, `:=`(Ftype=NULL, Page=NULL, PtypeID=NULL, Result=NULL, Units=NULL)]
loinc <- read_delim("data/hf_d_lab_procedure.tsv", "\t", col_types=cols(.default=col_character()))
setDT(loinc)
lab_counts <- lab[, .(N_lab = .N, N_pt=uniqueN(SK)), by="lab_procedure_id"]
lab_counts <- merge(lab_counts, loinc, by="lab_procedure_id", all.x=T, all.y=F)
setorder(lab_counts, -N_lab)
knitr::kable(lab_counts[1:20, .(lab_procedure_name, lab_super_group, N_lab, N_pt)])
```



## Facts: Medications

```{r warning=FALSE}
med <- read_delim("data/hf_neph_cohort_f_medication.tsv", ",", col_types=cols(.default=col_character(), Date=col_date()))
setnames(med, old="FID", new="medication_id")
setDT(med)
med[, `:=`(Ftype=NULL, Page=NULL, PtypeID=NULL)]
ndc <- read_delim("data/hf_d_medication.tsv", "\t", col_types=cols(.default=col_character()))
setDT(ndc)
med_counts <- med[, .(N_med = .N, N_pt=uniqueN(SK)), by="medication_id"]
med_counts <- merge(med_counts, ndc, by="medication_id", all.x=T, all.y=F)
setorder(med_counts, -N_med)
knitr::kable(med_counts[1:50, .(ndc_code, brand_name, generic_name, route_description, dose_form_description, N_med, N_pt)])
```

